ICT and Future Productivity:\\Evidence and Theory of a GPT\thanks{Correspondence: Department of Economics, Boston College, 140 Commonwealth Avenue, Chestnut Hill, MA 02467. Email: brianti@bc.edu (Marco Brianti) and gati@bc.edu (Laura G\'ati).}

### 2018-07-12 Latex/Draft_Summer2018/Summer2018.tex
more editing

The model presented has several common features with the two-sector environment of \cite{greenwood1997long} that design a two-sector economy where one sector displays a faster technology than the other. One different is that we focus specifically on ICT capital sector, rather than equipment, as the sector with faster technological improvements. This assumption will be necessary if we want to fit the fact that relative prices between the ICT investment and overall economy display a downward sloping trend.\footnote{In addition, we also rely on the results of \cite{whelan2003two} and \cite{oulton2007investment} to correctly define productivity and output share in our setting.}

@@

Consider an economy inhabited by a representative agent which maximizes the expect values of his lifetime utility as given by
\begin{eqnarray}\label{equation:utilFunc}
\E \Bigg[  \sum_{t=0}^{\infty} \beta^t U(c_t,l_t)  \Bigg]
\end{eqnarray}
with 
\begin{eqnarray*}
U(c_t,l_t) = \log(c_t) - \frac{1}{1+\frac{1}{\chi}} l_t^{1+\frac{1}{\chi}}
\end{eqnarray*}
where $c_t$ and $l_t$ represent consumption and labor at time $t$.

@@

Instead of working with the aggregate production function, assume that the production side of the final-good sector consist of a set $[0,1]$ of firms. In each period, final-good firm $j \in [0,1]$ produces $y^c_t(j)$ using the services of labor $l_{t}$, and two types of capital: hard capital, $k^c_{t}$ and ICT capital, $k^i_{t}$. In addition, production function is augmented with (i) neutral productivity common across sectors, (ii) sector-specific productivity, and (iii) a positive spillover effect related to the diffusion of ICT capital in the whole economy. Production of final output of firm $j$ is undertaken in accordance with
\begin{eqnarray}\label{equation:production_FINAL}
y^c_t(j) = A^c_t \ \big( k^c_{t}(j) \big)^a \ \big( k^i_{t}(j) \big)^b \ \big( l_{t}(j) \big)^{1-a-b}, \ \ 0 < a,b < 1
\end{eqnarray}
where total factor productivity is equal to $ A_t^c = \eta_t \ \theta^c_t \ (k^i_{t})^{\gamma} $. $\eta_t$ is the neutral technological component which moves along the following deterministic trend
\begin{eqnarray}\label{equation:neutral_tech_process}
\eta_t = (\Gamma^{\eta})^t e^{\varepsilon_t},  
\end{eqnarray} 
\begin{eqnarray}\label{equation:neutral_tech_shock}
 \varepsilon_t = \rho_{\nu} \varepsilon_{t-1} + \nu_t
\end{eqnarray} 
where $\rho_{\nu} \in (0,1)$ and $\nu_t \sim N(0,\sigma_{\nu})$ represents a neutral technology shock. $\theta^c_t$ is the sector-specific technological component,  which moves along the following deterministic trend
\begin{eqnarray}
\theta_t^c = (\Gamma^{c})^t.
\end{eqnarray} 
Then, $\gamma$ represents the parameter that governs the size of the spillover effect of ICT capital in the production function. Moreover, $k^c_{t}(j)$ is the part of hard capital $k^c_t$ used by firm $j$ in oder to produce final-good output. Similarly, $k^i_{t}(j)$ and $l_{t}(j)$ are the parts of ICT capital $k^i_t$ and labor $l_t$ used by firm $j$ in oder to produce final-good output, respectively.

Since all the final-good producers are identical ex ante and ex post, the following conditions hold,
\begin{eqnarray}\label{equation:aggregation_hard_final}
\int_0^1 k^c_{t}(j) dj = k^c_{1,t},
\end{eqnarray}
\begin{eqnarray}\label{equation:aggregation_ICT_final}
\int_0^1 k^i_{t}(j) dj = k^i_{1,t}, 
\end{eqnarray}
\begin{eqnarray}\label{equation:aggregation_labor_final}
\int_0^1 l_{t}(j) dj = l_{1,t},
\end{eqnarray}
where $k^c_{1,t}$ is the aggregate part of hard capital $k^c_t$ devoted to the final-sector, i.e. sector 1. Moreover, $k^i_{1,t}$ and $l_{1,t}$ are the aggregate parts of ICT capital $k^i_{t}$ and labor $l_{t}$ devoted to the final-good sector, respectively.

@@

Instead of working with the aggregate production function, assume that the production side of the ICT-good sector consist of a set $[0,1]$ of firms. Similar to final-output production function, ICT-good firm $q \in [0,1]$ produces $y^c_t(q)$ using the services of labor $l_{t}$, and two types of capital: hard capital, $k^c_{t}$ and ICT capital, $k^i_{t}$. In addition, production function is augmented with (i) neutral productivity common across sectors, (ii) sector-specific productivity, and (iii) a positive spillover effect related to the diffusion of ICT capital in the whole economy. Production of ICT output of firm $q$ is undertaken in accordance with
\begin{eqnarray}\label{equation:productionICT}
y^c_t(q) = A^c_t \ \big( k^c_{t}(q) \big)^a \ \big( k^i_{t}(q) \big)^b \ \big( l_{t}(q) \big)^{1-a-b}, \ \ 0 < a,b < 1
\end{eqnarray}
where total factor productivity is equal to $ A_t^c = \eta_t \ \theta^i_t \ (k^i_{t})^{\gamma} $. $\eta_t$ is the same neutral technological component which follows \ref{equation:neutral_tech_process} and \ref{equation:neutral_tech_shock}, $\theta^i_t$ is the sector-specific technological component which moves along the following deterministic trend
\begin{eqnarray}\label{equation:ICT_tech_process}
\theta^i_t = (\Gamma^{i})^t e^{\zeta_t},  
\end{eqnarray} 
\begin{eqnarray}\label{equation:ICT_tech_shock}
\zeta_t = \rho_{\iota} \zeta_{t-1} + \iota_t
\end{eqnarray} 
where $\rho_{\iota} \in (0,1)$ and $\iota_t \sim N(0,\sigma_{\iota})$ represents an ICT technology shock.
Then, $\gamma$ represents the same parameter that governs the size of the spillover effect of ICT capital in the production function. Moreover, $k^c_{t}(q)$ is the part of hard capital $k^c_t$ used by firm $q$ in oder to produce ICT goods. Similarly, $k^i_{t}(q)$ and $l_{t}(q)$ are the parts of ICT capital $k^i_t$ and labor $l_t$ used by firm $q$ in oder to produce ICT goods, respectively.

Similarly, since all the final-good producers are identical ex ante and ex post, the following conditions hold,
\begin{eqnarray}\label{equation:aggregation_hard_ICTsec}
\int_0^1 k^c_{t}(q) dq = k^c_{2,t},
\end{eqnarray}
\begin{eqnarray}\label{equation:aggregation_ICT_ICTsec}
\int_0^1 k^i_{t}(q) dq = k^i_{2,t}, 
\end{eqnarray}
\begin{eqnarray}\label{equation:aggregation_labor_ICTsec}
\int_0^1 l_{t}(q) dq = l_{2,t},
\end{eqnarray}
where $k^c_{2,t}$ is the aggregate part of hard capital $k^c_t$ devoted to the ICT-sector, i.e. sector 2. Moreover, $k^i_{2,t}$ and $l_{2,t}$ are the aggregate parts of ICT capital $k^i_{t}$ and labor $l_{t}$ devoted to the ICT-good sector, respectively.

Aggregate ICT-good output is defined as
\begin{eqnarray}\label{equation:production_ICT_aggregate}
	y^i_t = A^c_t \ ( k^c_{2,t} )^a \ ( k^i_{2,t} )^b \ ( l_{2,t} )^{1-a-b}, \ \ 0 < a,b < 1
\end{eqnarray}
and it can be only investment in ICT capital, $i^i_t$,
\begin{eqnarray}\label{equation:resourceICT}
y^i_t = i^i_t
\end{eqnarray}
Notice that elements in the equation above are normalized such that both output and investment are measured in the same units. Moreover, ICT goods can be produced from ICT output on a one-to-one basis. The stock of ICT capital evolves according to
\begin{eqnarray}\label{equation:LOM_ICT}
k^i_{t+1} = (1 - \delta^i) k^i_t + i^i_t
\end{eqnarray}
where $0 < \delta^i < 1$ represents the period depreciation of ICT capital.

@@

Since we are assuming our economy to be closed, input demands must equal supplies,
\begin{eqnarray}\label{equation:marketclearingHard}
k^c_t = k^c_{1,t} + k^c_{2,t},
\end{eqnarray}
\begin{eqnarray}\label{equation:marketclearingICT}
k^i_t = k^i_{1,t} + k^i_{2,t}, 
\end{eqnarray}
\begin{eqnarray}\label{equation:marketclearingLabor}
l_t = l_{1,t} + l_{2,t}.
\end{eqnarray}
Taking as given Equations \ref{equation:production_FINAL}-\ref{equation:marketclearingLabor} and assuming that factor markets are perfective competitive, representative household maximizes intertemporal utility function under the following constraint,
\begin{eqnarray}\label{equation:resourceHH}
	c_t + k^c_{t+1} +p_t k^i_{t+1} = \bigg( 1 - \delta^c + r^c_t \bigg) k^c_t + \bigg( 1 - \delta^i + \frac{r^i_t}{p_t} \bigg) p_t k^i_t + w_t l_t	
\end{eqnarray}
where $p_t$ represents the price of ICT capital expressed in consumption units. Since price of consumption is normalized to be one then $p_t$ also represents the relative price between final output and ICT output.

>>>>

The model presented has several common features with the two-sector environment of \cite{greenwood1997long} where one sector displays a faster technology than the other. One difference is that we focus specifically on the ICT capital sector, rather than equipment, as the sector with faster technological improvement. This assumption will be necessary if we want to fit the fact that relative prices between ICT investment and the overall economy display a downward sloping trend.\footnote{In addition, we also rely on the results of \cite{whelan2003two} and \cite{oulton2007investment} to correctly define productivity and output share in our setting.} A second difference is that  we consider two production functions - one for each sector. As \cite{oulton2007investment} shows, the two formulations are isomorphic, so our modeling choice has no implication for the resulting dynamics. The reason we choose this approach is because it allows for definitions of GDP and TFP in the model that are consistent with the way these variables are constructed in the national income accounts.

@@

Consider an economy inhabited by a representative agent who maximizes the expected values of his lifetime utility as given by
\begin{eqnarray}\label{equation:utilFunc}
\E \Bigg[  \sum_{t=0}^{\infty} \beta^t U(c_t,l_t)  \Bigg]
\end{eqnarray}
with 
\begin{eqnarray*}
U(c_t,l_t) = \log(c_t) - \frac{\chi}{2} l_t^{2}
\end{eqnarray*}
%\begin{eqnarray*}
%U(c_t,l_t) = \log(c_t) - \frac{1}{1+\frac{1}{\chi}} l_t^{1+\frac{1}{\chi}}
%\end{eqnarray*}
where $c_t$ and $l_t$ represent consumption and labor at time $t$.

@@

The production side of the final-good sector consists of a set $[0,1]$ of firms. In each period, final-good firm $j \in [0,1]$ produces $y^c_t(j)$ using the services of labor $l_{t}$, and two types of capital: hard capital, $k^c_{t}$ and ICT capital, $k^i_{t}$. In addition, the production function is augmented with (i) neutral productivity common across sectors, (ii) sector-specific productivity, and (iii) a positive spillover effect related to the diffusion of ICT capital in the whole economy. Production of final output of firm $j$ is undertaken in accordance with
\begin{eqnarray}\label{equation:production_FINAL}
y^c_t(j) = A^c_t \ \big( k^c_{t}(j) \big)^a \ \big( k^i_{t}(j) \big)^b \ \big( l_{t}(j) \big)^{1-a-b}, \ \ 0 < a,b < 1
\end{eqnarray}
where total factor productivity is equal to $ A_t^c = \eta_t \ \theta^c_t \ (k^i_{t})^{\gamma} $. $\eta_t$ is the neutral technological component which moves along the following deterministic trend
\begin{eqnarray}\label{equation:neutral_tech_process}
\eta_t = (\Gamma^{\eta})^t e^{\varepsilon_t},  
\end{eqnarray} 
\begin{eqnarray}\label{equation:neutral_tech_shock}
 \varepsilon_t = \rho_{\nu} \varepsilon_{t-1} + \nu_t
\end{eqnarray} 
where $\rho_{\nu} \in (0,1)$ and $\nu_t \sim N(0,\sigma_{\nu})$ represents a neutral technology shock. $\theta^c_t$ is the sector-specific technological component,  which moves along the following deterministic trend
\begin{eqnarray}
\theta_t^c = (\Gamma^{c})^t.
\end{eqnarray} 
Then, $\gamma$ represents the parameter that governs the size of the spillover effect of ICT capital in the production function. Moreover, $k^c_{t}(j)$ is the part of hard capital $k^c_t$ used by firm $j$ in oder to produce final-good output. Similarly, $k^i_{t}(j)$ and $l_{t}(j)$ are the parts of ICT capital $k^i_t$ and labor $l_t$ used by firm $j$ in oder to produce final-good output, respectively.

Since all the final-good producers are identical, the following conditions hold,
\begin{eqnarray}\label{equation:aggregation_hard_final}
\int_0^1 k^c_{t}(j) dj = k^c_{1,t},
\end{eqnarray}
\begin{eqnarray}\label{equation:aggregation_ICT_final}
\int_0^1 k^i_{t}(j) dj = k^i_{1,t}, 
\end{eqnarray}
\begin{eqnarray}\label{equation:aggregation_labor_final}
\int_0^1 l_{t}(j) dj = l_{1,t},
\end{eqnarray}
where $k^c_{1,t}$ is the aggregate part of hard capital $k^c_t$ devoted to the final-sector, i.e. sector 1. Moreover, $k^i_{1,t}$ and $l_{1,t}$ are the aggregate parts of ICT capital $k^i_{t}$ and labor $l_{t}$ devoted to the final-good sector, respectively.

@@

The production side of the ICT-good sector is symmetric to the first sector. It consists of a set $[0,1]$ of firms. Similar to final-output production function, ICT-good firm $q \in [0,1]$ produces $y^c_t(q)$ using the services of labor $l_{t}$, and two types of capital: hard capital, $k^c_{t}$ and ICT capital, $k^i_{t}$. In addition, production function is augmented with the same components as the other sector: (i) neutral productivity common across sectors, (ii) sector-specific productivity, and (iii) a positive spillover effect related to the diffusion of ICT capital in the whole economy. Production of ICT output of firm $q$ is undertaken in accordance with
\begin{eqnarray}\label{equation:productionICT}
y^i_t(q) = A^i_t \ \big( k^c_{t}(q) \big)^a \ \big( k^i_{t}(q) \big)^b \ \big( l_{t}(q) \big)^{1-a-b}, \ \ 0 < a,b < 1
\end{eqnarray}
where total factor productivity is equal to $ A_t^i = \eta_t \ \theta^i_t \ (k^i_{t})^{\gamma} $. $\eta_t$ is the same neutral technological component which follows \ref{equation:neutral_tech_process} and \ref{equation:neutral_tech_shock}, $\theta^i_t$ is the sector-specific technological component which moves along the following deterministic trend
\begin{eqnarray}\label{equation:ICT_tech_process}
\theta^i_t = (\Gamma^{i})^t e^{\zeta_t},  
\end{eqnarray} 
\begin{eqnarray}\label{equation:ICT_tech_shock}
\zeta_t = \rho_{\iota} \zeta_{t-1} + \iota_t
\end{eqnarray} 
where $\rho_{\iota} \in (0,1)$ and $\iota_t \sim N(0,\sigma_{\iota})$ represents an ICT technology shock.
Then, $\gamma$ represents the same parameter that governs the size of the spillover effect of ICT capital in the production function. Moreover, $k^c_{t}(q)$ is the part of hard capital $k^c_t$ used by firm $q$ in oder to produce ICT goods. Similarly, $k^i_{t}(q)$ and $l_{t}(q)$ are the parts of ICT capital $k^i_t$ and labor $l_t$ used by firm $q$ in oder to produce ICT goods, respectively.

Similarly, since all the final-good producers are identical, the following conditions hold,
\begin{eqnarray}\label{equation:aggregation_hard_ICTsec}
\int_0^1 k^c_{t}(q) dq = k^c_{2,t},
\end{eqnarray}
\begin{eqnarray}\label{equation:aggregation_ICT_ICTsec}
\int_0^1 k^i_{t}(q) dq = k^i_{2,t}, 
\end{eqnarray}
\begin{eqnarray}\label{equation:aggregation_labor_ICTsec}
\int_0^1 l_{t}(q) dq = l_{2,t},
\end{eqnarray}
where $k^c_{2,t}$ is the aggregate part of hard capital $k^c_t$ devoted to the ICT-sector, i.e. sector 2. Moreover, $k^i_{2,t}$ and $l_{2,t}$ are the aggregate parts of ICT capital $k^i_{t}$ and labor $l_{t}$ devoted to the ICT-good sector, respectively.

Aggregate ICT-good output is defined as
\begin{eqnarray}\label{equation:production_ICT_aggregate}
	y^i_t = A^c_t \ ( k^c_{2,t} )^a \ ( k^i_{2,t} )^b \ ( l_{2,t} )^{1-a-b}, \ \ 0 < a,b < 1
\end{eqnarray}
and it can be only investment in ICT capital, $i^i_t$,
\begin{eqnarray}\label{equation:resourceICT}
y^i_t = i^i_t
\end{eqnarray}
Also here the elements in the equation above are normalized such that both output and investment are measured in the same units. Moreover, ICT output is converted to ICT goods on a one-to-one basis. The stock of ICT capital evolves according to
\begin{eqnarray}\label{equation:LOM_ICT}
k^i_{t+1} = (1 - \delta^i) k^i_t + i^i_t
\end{eqnarray}
where $0 < \delta^i < 1$ represents the period depreciation of ICT capital.

@@

Since the economy is closed, input demands must equal supplies,
\begin{eqnarray}\label{equation:marketclearingHard}
k^c_t = k^c_{1,t} + k^c_{2,t},
\end{eqnarray}
\begin{eqnarray}\label{equation:marketclearingICT}
k^i_t = k^i_{1,t} + k^i_{2,t}, 
\end{eqnarray}
\begin{eqnarray}\label{equation:marketclearingLabor}
l_t = l_{1,t} + l_{2,t}.
\end{eqnarray}
Taking as given Equations \ref{equation:production_FINAL}-\ref{equation:marketclearingLabor} and assuming that factor markets are perfective competitive, the representative household maximizes intertemporal utility under the following constraint,
\begin{eqnarray}\label{equation:resourceHH}
	c_t + k^c_{t+1} +p_t k^i_{t+1} = \bigg( 1 - \delta^c + r^c_t \bigg) k^c_t + \bigg( 1 - \delta^i + \frac{r^i_t}{p_t} \bigg) p_t k^i_t + w_t l_t	
\end{eqnarray}
where $p_t$ represents the price of ICT capital expressed in consumption units. Since the price of consumption is normalized to be 1, $p_t$ also represents the relative price between final output and ICT output.
